---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r, echo = FALSE}
# Turn off printing in scientific notation.
inline_hook <- function(x) {
  if (is.numeric(x)) {
    format(x, digits = 2)
  } else x
}
knitr::knit_hooks$set(inline = inline_hook)
```

Load R packages:
```{r}
library(tidyverse)
library(lubridate)
library(xtable)
library(impute)
```

## Loading and preprocessing the data

### 1. Load the data (i.e. read.csv())).
Unzip and load the file.  
```{r}
if(!file.exists("activity.csv")) {
  unzip("activity.zip") }
data <- read.csv("activity.csv")
```  

### 2. Process/transform the data (if necessary) into a format suitable for your analysis.  

```{r}
data$steps <- as.numeric(data$steps)
```

## What is mean total number of steps taken per day?  

For this part of the assignment, ignore the missing values in the dataset.
```{r}
dataComplete <- drop_na(data)
```

### 1. Calculate the total number of steps taken per day.  
```{r, results = "asis"}
stepsTotalByDay <- group_by(dataComplete, date) %>%
        summarize(stepsTotal = sum(steps, na.rm = TRUE))
print(xtable(stepsTotalByDay), type = "html", include.rownames = FALSE)
```  
  
  
### 2. Make a histogram of the total number of steps taken each day.  
```{r}
ggplot(data = stepsTotalByDay, aes(stepsTotal)) +
        geom_histogram(binwidth = 1000) +
        labs(x = "Total number of steps taken per day")
```

### 3. Calculate and report the mean and median of the total number of steps taken per day.    
```{r}
stepsTotalByDayMean <- mean(stepsTotalByDay$stepsTotal)
stepsTotalByDayMedian <- median(stepsTotalByDay$stepsTotal)
```

The mean of the total number of steps taken per day is `r stepsTotalByDayMean` steps.
The median of the total number of steps taken per day is `r stepsTotalByDayMedian` steps.

## What is the average daily activity pattern?

### 1. Make a time series plot (i.e. type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).  

Time series plot of average number of steps taken in 5-minute intervals, averaged across all days

```{r}
stepsAverageByInterval <- group_by(data, interval) %>%
        summarize(stepsAverage = mean(steps, na.rm = TRUE))

ggplot(data = stepsAverageByInterval, aes(x = interval, y = stepsAverage)) +
        geom_line() +
        labs(x = "5-minute interval",
             y = "Average number of steps")
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxInterval <- filter(stepsAverageByInterval, stepsAverage == max(stepsAverage))
maxIntervalNumber <- maxInterval[[1]]
maxIntervalSteps <- round(maxInterval[[2]], 0)
```

The 5-minute interval with maximum average number of steps is interval `r maxIntervalNumber`, which has an average of `r maxIntervalSteps` steps.  

## Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)  
```{r}
missingValues <- sum(is.na(data$steps))
totalValues <- nrow(data)

```

The total number of missing values in the data set is `r missingValues` out of 
`r totalValues` total values.  

### 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.  

Impute missing data using impute.knn function in impute package, and create a 
new dataset (dataImputed) with missing values imputed.
```{r}
dataMatrix <- as.matrix(data[ ,c(1,3)])
dataImputedTmp <- impute.knn(dataMatrix, rng.seed = 1)
```

### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.  
```{r}
dataImputed <- as.data.frame(dataImputedTmp$data)
dataImputed <- mutate(dataImputed, date = data$date)
```

### 4. Make a histogram of the total number of steps taken each day. Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?  

Make a histogram of the total number of steps taken each day.
```{r}
stepsTotalByDayImputed <- group_by(dataImputed, date) %>%
        summarize(stepsTotal = sum(steps, na.rm = TRUE))
ggplot(data = stepsTotalByDayImputed, aes(stepsTotal)) +
        geom_histogram(binwidth = 1000) +
        labs(x = "Total number of steps taken per day")
```

Calculate and report the mean and median total number of steps taken per day.
```{r}
stepsTotalByDayImputedMean <- mean(stepsTotalByDayImputed$stepsTotal)
stepsTotalByDayImputedMedian <- median(stepsTotalByDayImputed$stepsTotal)
```

In the dataset including the imputed data, the mean of the total number of steps taken per day is `r stepsTotalByDayImputedMean` steps.
In the dataset including the imputed data, the median of the total number of steps taken per day is `r stepsTotalByDayImputedMedian` steps.

Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
stepsMeanDiff = stepsTotalByDayImputedMean - stepsTotalByDayMean
stepsMedianDiff = stepsTotalByDayImputedMedian - stepsTotalByDayMedian
```
The mean of the total number of steps taken per day is `r stepsTotalByDayMean` and `r stepsTotalByDayImputedMean` from the original dataset and dataset with imputed values, respectively. This is a difference of `r stepsMeanDiff`.  

The median of the total number of steps taken per day is `r stepsTotalByDayMedian` and `r stepsTotalByDayImputedMedian` from the original dataset and dataset with imputed values, respectively. This is a difference of `r stepsMedianDiff`.  

## Are there differences in activity patterns between weekdays and weekends?

### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
dataImputed <- dataImputed %>%
        mutate(day = wday(ymd(dataImputed$date), label = TRUE, abbr = FALSE)) %>%
        mutate(weekend = ifelse(day == "Saturday" | day == "Sunday", 
                                "weekend", "weekday"))
```

### 2. Make a panel plot containing a time series plot (i.e. type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
stepsAverageByIntervalImputed <- group_by(dataImputed, weekend, interval) %>%
        summarize(stepsAverage = mean(steps, na.rm = TRUE))

ggplot(data = stepsAverageByIntervalImputed, 
       aes(x = interval, y = stepsAverage)) +
        geom_line() +
        facet_wrap(vars(weekend), nrow = 2) + 
        labs(x = "5-minute interval",
             y = "Average number of steps")
```

